The universal property of infinite direct sums in C$^*$-categories and W$^*$-categories

### 2019-07-10 direct_sums.tex
acknowlegements
updates

\thanks{\textit{Acknowledgements.} We thank Robert Furber for copious help and discussion. Part of this work was conducted while the first author was at the Max Planck Institute for Mathematics in the Sciences in Leipzig, Germany.}

@@

\subsection*{C*-algebras and W*-algebras}  We give a very brief recap of C*-algebras and W*-algebras for the benefit of readers who have a background in category theory but not in operator algebras, while referring to the literature for the details~\cite{sakai}.

\emph{C$^*$-algebras} are abstract versions of algebras of operators on a Hilbert space which are closed in the operator norm and closed under taking adjoint operators, where taking adjoints is encoded in the ``star'' operation $*$. Similarly, \emph{W$^*$-algebras} are abstract versions of algebras of operators on a Hilbert space which are in addition closed under limits with respect to pointwise convergence (strong operator topology). In fact, every abstract C$^*$-algebra or W$^*$-algebra is isomorphic to one having such a concrete representation. For example, every $*$-closed subalgebra of the algebra of complex~$n\times n$-matrices is a C$^*$-algebra and a~W$^*$-algebra. Moreover, every commutative unital C$^*$-algebra is isomorphic to~$C(X)$ for some compact Hausdorff space~$X$. And every commutative W$^*$-algebra is isomorphic to~$L^\infty(X,\mu)$ for some complete finite measure space~$(X,\mu)$.

@@

A category can be thought of as a many-object version of a monoid. This manifests itself in the fact that for any category $\cat{C}$, every endomorphism homset $\cat{C}(A,A)$ is a monoid. In the same way, a C$^*$-category is a many-object version of a unital C$^*$-algebra. Indeed in a~C$^*$-category~$\cat{C}$, every homset~$\cat{C}(A,A)$ will be a unital~C$^*$-algebra. Similarly, a W$^*$-category $\cat{C}$ is a many-object version of a W$^*$-algebra, and every homset~$\cat{C}(A,A)$ indeed carries the structure of a~W$^*$-algebra.

@@

\begin{defn}
    Let~$\cat{C}$ be a category together with an endofunctor~$*\,\colon \cat{C} \to \cat{C}^\op$,
    whose action on arrows we denote by~$a\mapsto a^*$. The pair $(\cat{C},*)$ is called a \emph{complex~$*$-category} if the following conditions hold.
\begin{enumerate}
    \item $\cat{C}$ is enriched over complex vector spaces, meaning that:
	    \begin{enumerate}
		    \item every homset~$\cat{C}(A,B)$ is a complex vector space,
		    \item composition of morphisms is bilinear.
	   \end{enumerate}
    \item The contravariant endofunctor~$*$ is
	    \begin{enumerate}
		    \item identity-on-objects: $a^* : B \to A$ for $a : A \to B$,
		    \item involutive: $a^{**} = a$,
		    \item conjugate linear:
			    \[
				    (\lambda a)^* = \overline{\lambda} a^*,\qquad (a+b)^* = a^* + b^*
			    \]
			    for morphisms $a, b : A \to B$ and $\lambda\in\C$.
	    \end{enumerate}
    \item For every morphism~$a : A \to B$,
	    \begin{enumerate}
		    \item There is $b :  A \to A$ with~$a^*a=b^*b$,
		    \item $a^*a=0$ implies $a = 0$.
	    \end{enumerate}
\end{enumerate}
\end{defn}

@@

\begin{defn}
A complex $*$-category $(\cat{C},*)$ is called a
\begin{enumerate}
	\item \emph{normed $*$-category} if each homset~$\cat{C}(A,B)$ is a normed space such that
	    \[
		    \| ab \| \leq \|a\| \, \|b\|
	    \]
                for any~$a\colon B \to C$ and~$b\colon A \to B$.
	\item\label{Cstarcat}
    \emph{C$^*$-category} if moreover every homset is norm-complete, and the C$^*$-identity
    \[
	\|a \|^2 = \|a^*a\|
     \]
	holds for every morphism $a$.
\item\label{Wstarcat}
    \emph{W$^*$-category} if moreover
        where every homset has a Banach space predual.

	Equivalently~\cite[Prop.~2.15]{wstarcat}, a C$^*$-category~$\cat{C}$
            is a~W$^*$-category
            if every~$\cat{C}(A,A)$ is a W$^*$-algebra
            and~$\cat{C}(A,B)$ is self-dual
            as a Hilbert~$(A,A)$-module
            with inner product~$\langle a,b\rangle \equiv a^*b$.
\end{enumerate}
\end{defn}

@@

\begin{ex}
\label{rep_ex}
If $A$ is a C$^*$-algebra, then the category of non-degenerate representations $\pi : A \to \B{\H}$ on Hilbert spaces $\H$ is a W$^*$-category. The morphisms $(\H_1,\pi_1) \to (\H_2,\pi_2)$ are exactly the \emph{intertwiners}, i.e.~the bounded linear maps $f : \H_1 \to \H_2$ such that the diagram
\[\begin{tikzcd}
	\H_1 \ar{r}{f} \ar[swap]{d}{\pi_1(a)} & \H_2 \ar{d}{\pi_2(a)} \\
	\H_1 \ar{r}{f} & \H_2
\end{tikzcd}\]
commutes for all $a\in A$. It is well-known that this is a W$^*$-category~\cite[Example~2.4]{wstarcat}, where the Banach space structure on each hom-set is the usual one on a space of operators using the operator norm, and the involution is given by the usual Hermitian adjoint. We denote this W$^*$-category by $\Rep{A}$. In the most basic case $A = \C$, we get $\Rep{\C} \cong \Hilb$, the category of Hilbert spaces.
\end{ex}

@@

We again let $\cat{C}$ be a C$^*$-category. Let $I$ be a finite or infinite index set. For $(A_i)_{i\in I}$ a family of objects in $\cat{C}$, we consider the functor
\[
	\bigoplus_i \cat{C}(A_i,-) \: : \: \cat{C} \to \Ban
\]
given by associating to every object $B$ the set
\beq
\label{AiB}
	\bigoplus_i \cat{C}(A_i , B) \ := \ \left\{ (f_i : A_i \to B)_{i\in I} \Biggm\vert \sum_i f_i f_i^* < \infty \right\}.
\eeq
Here, $\sum_i f_i f_i^*$ can be interpreted as converging in the ultraweak topology\footnote{For positive elements~$(x_i)_{i \in I}$ of a W$^*$-algebra, convergence and value of the sum~$\sum_{i \in I} x_i$ coincides in the ultraweak and ultrastrong topology and, in fact, equals the supremum of the partial sums.} in the case where $\cat{C}(B,B)$ is a W$^*$-algebra. If this is not the case, what we mean by the condition $\sum_i f_i f_i^* < \infty$ is that there is a fixed element which upper bounds all partial sums $\sum_{i\in S} f_i f_i^*$ for finite $S\subseteq I$, or equivalently that the sum converges ultraweakly in the double dual W$^*$-algebra $\cat{C}(B,B)^{**}$.

The set~\eqref{AiB} is a vector space under componentwise addition and scalar multiplication, where closure under addition follows from $f_i g^*_i + g_i f^*_i \leq f_i f^*_i + g_i g^*_i$, which holds because of $(f_i - g_i) (f^*_i - g^*_i) \geq 0$. We furthermore consider this vector space as a normed space under the norm
\[
	\| (f_i)_{i \in I} \| := \left\| \sum_i f_i f_i^* \right\|^{1/2},
\]
where the triangle inequality again follows from $f_i g^*_i + g_i f^*_i \leq f_i f^*_i + g_i g^*_i$. We interpret the sum as before either in terms of a supremum over the norm of finite partial sums, or as converging in the double dual $\cat{C}(B,B)^{**}$.

@@

\begin{proof}
Suppose that we are given a Cauchy sequence of families $(f^{(n)})_{n\in\N} = ( ( f_i^{(n)} )_{i\in I} )_{n\in\N}$. For given $j$, the projection map
\[
	\bigoplus_i \cat{C}(A_i,B) \longrightarrow \cat{C}(A_j,B)	
\]
is bounded, and therefore $( f_j^{(n)} )_{n\in\N}$ is Cauchy as well; let us denote the limit by $f_j : A_j \to B$. We will now show that $f \in \bigoplus_i \cat{C}(A_i,B)$ for $f := (f_j)_{j\in I}$, which amounts to verifying the relevant square summability. The Cauchy assumption implies that the sequence $(f^{(n)})_{n\in\N}$ is uniformly bounded by some constant $C$,
\[
	\left\| \sum_i f_i^{(n)} (f_i^{(n)})^* \right\|^{1/2} \leq C.
\]
Restricting to a finite partial sum over $S\subseteq I$ and taking $n\to\infty$ shows that
\[
	\left\| \sum_{i\in S} f_i f_i^* \right\|^{1/2} \leq C
\]
as well. Since $S$ was arbitrary, we conclude that $\| f \| \leq C$.

The convergence $f_i \to f$ works similarly. By assumption, for every $\eps > 0$ we have $\left\| f^{(n)} - f^{(m)} \right\| \leq \eps$ for almost all $n,m\in\N$. In particular, we know that for any finite $S \subseteq I$,
\[
	\left\| \sum_{i\in S} (f_i^{(n)} - f_i^{(m)})^* (f_i^{(n)} - f_i^{(m)}) \right\|^{1/2} \leq \eps
\]
for almost all $n,m\in\N$. Taking $m\to\infty$ gives
\[
	\left\| \sum_{i\in S} (f_i^{(n)} - f_i)^* (f_i^{(n)} - f_i) \right\|^{1/2} \leq \eps.
\]
Since $S$ was arbitrary, we conclude $\| f^{(n)} - f \| \leq \eps$ for almost all $n$.
\end{proof}

>>>>

\thanks{\textit{Acknowledgements.} We thank Robert Furber for copious help and discussion, as well as Chris Heunen and Martti Korvanen for discussion. Part of this work was conducted while the first author was at the Max Planck Institute for Mathematics in the Sciences in Leipzig, Germany.}

@@

\subsection*{C*-algebras and W*-algebras}  We give a brief recap of C*-algebras and W*-algebras for the benefit of readers who have a background in category theory but not in operator algebras, while referring to the literature for the details~\cite{sakai,kr,bram}.

\emph{C$^*$-algebras} are abstract versions of algebras of operators
on a Hilbert space which are closed in the operator norm and closed
under taking adjoint operators, where taking adjoints is encoded
in the ``star'' operation $*$. Similarly, \emph{W$^*$-algebras} are
abstract versions of algebras of operators on a Hilbert space which
are in addition closed under limits with respect to the strong
operator topology, a sort of pointwise converge.\footnote{A W$^*$-algebra
    of operators is usually called a von Neumann algebra.
    However, our references \cite{bas,bram} use the terms W$^*$-algebra and
        von Neumann algebra interchangeably.}
In fact, every abstract C$^*$-algebra or W$^*$-algebra is isomorphic
to one having such a concrete representation~\cite[\S30\textsubscript{XIV}, \S48\textsubscript{VII}]{bram}.
For example, every $*$-closed subalgebra
of the algebra of complex~$n\times n$-matrices is a C$^*$-algebra
and a~W$^*$-algebra. Moreover, every commutative unital C$^*$-algebra
is isomorphic to~$C(X)$ for some compact Hausdorff space~$X$~\cite[\S27]{bram}. And
every commutative W$^*$-algebra is isomorphic to~$L^\infty(X,\mu)$
for some complete finite measure space~$(X,\mu)$~\cite[\S53\textsubscript{XI}]{bram}.

@@

A category can be thought of as a many-object version of a monoid.
This manifests itself in the fact that for any category $\cat{C}$,
every endomorphism homset $\cat{C}(A,A)$ is a monoid. In the same
way, a C$^*$-category is a many-object version of a unital
C$^*$-algebra. Indeed in a~C$^*$-category~$\cat{C}$, every
homset~$\cat{C}(A,A)$ will be a unital~C$^*$-algebra. Similarly, a
W$^*$-category $\cat{C}$ is a many-object version of a W$^*$-algebra,
and in a W$^*$-category every homset~$\cat{C}(A,A)$ carries the structure of
a~W$^*$-algebra.

@@

\begin{defn}
    Let~$\cat{C}$ be a category together with an endofunctor~$*\,\colon \cat{C} \to \cat{C}^\op$,
    whose action on arrows we denote by~$a\mapsto a^*$. The pair $(\cat{C},*)$ is called a \emph{complex~$*$-category} if the following conditions hold.
\begin{enumerate}
    \item $\cat{C}$ is enriched over complex vector spaces, meaning that:
	    \begin{enumerate}
		    \item every homset~$\cat{C}(A,B)$ is a complex vector space,
		    \item composition of morphisms is bilinear.
	   \end{enumerate}
    \item The contravariant endofunctor~$*$ is
	    \begin{enumerate}
		    \item identity-on-objects: $a^* : B \to A$ for $a : A \to B$,
		    \item involutive: $a^{**} = a$,
		    \item conjugate linear:
			    \[
				    (\lambda a)^* \ =\  \overline{\lambda} a^*,\qquad (a+b)^* \ =\  a^* + b^*
			    \]
			    for morphisms $a, b : A \to B$ and $\lambda\in\C$.
	    \end{enumerate}
    \item For every morphism~$a : A \to B$,
	    \begin{enumerate}
		    \item there is a $b :  A \to A$ with~$a^*a=b^*b$,
		    \item $a^*a=0$ implies $a = 0$.
	    \end{enumerate}
\end{enumerate}
\end{defn}

@@

\begin{defn}
A complex $*$-category $(\cat{C},*)$ is called a
\begin{enumerate}
	\item \emph{normed $*$-category} if each homset~$\cat{C}(A,B)$ is a normed space such that
	    \[
		    \| ab \| \ \leq\  \|a\| \, \|b\|
	    \]
                for any~$a\colon B \to C$ and~$b\colon A \to B$.
	\item\label{Cstarcat}
    \emph{C$^*$-category} if moreover every homset is norm-complete, and the C$^*$-identity
    \[
	\|a \|^2\  =\  \|a^*a\|
     \]
	holds for every morphism $a$.
\item\label{Wstarcat}
    \emph{W$^*$-category} if moreover
        where every homset has a Banach space predual.

	Equivalently~\cite[Prop.~2.15]{wstarcat}, a C$^*$-category~$\cat{C}$
            is a~W$^*$-category
            if every~$\cat{C}(A,A)$ is a W$^*$-algebra
            and~$\cat{C}(A,B)$ is self dual
            as a Hilbert~$\cat{C}(A,A)$-module
            with inner product~$\langle a,b\rangle \equiv a^*b$.
\end{enumerate}
\end{defn}

@@

\begin{ex}
\label{rep_ex}
If $A$ is a C$^*$-algebra, then the category of non-degenerate
representations $\pi : A \to \B{\H}$ on Hilbert spaces $\H$ is a
W$^*$-category. The morphisms~$(\H_1,\pi_1) \to (\H_2,\pi_2)$ are
exactly the \emph{intertwiners}, i.e.~the bounded linear maps~$f :
\H_1 \to \H_2$ such that the diagram
\[\begin{tikzcd}
	\H_1 \ar{r}{f} \ar[swap]{d}{\pi_1(a)} & \H_2 \ar{d}{\pi_2(a)} \\
	\H_1 \ar{r}{f} & \H_2
\end{tikzcd}\]
commutes for all $a\in A$. It is well-known that this is a W$^*$-category~\cite[Example~2.4]{wstarcat}, where the Banach space structure on each hom-set is the usual one on a space of operators using the operator norm, and the involution is given by the usual Hermitian adjoint. We denote this W$^*$-category by $\Rep{A}$. In the most basic case $A = \C$, we get $\Rep{\C} \cong \Hilb$, the category of Hilbert spaces.
\end{ex}

@@

We again let $\cat{C}$ be a C$^*$-category. Let $I$ be a finite or infinite index set. For $(A_i)_{i\in I}$ a family of objects in $\cat{C}$, we consider the functor
\[
	\bigoplus_i \cat{C}(A_i,-) \colon \  \cat{C} \to \Ban
\]
given by associating to every object $B$ the set
\beq
\label{AiB}
	\bigoplus_i \cat{C}(A_i , B) \ := \ \Bigl\{ (f_i : A_i \to B)_{i\in I} \Bigm\vert\textstyle  \sum_i f_i f_i^* < \infty \Bigr\}.
\eeq
Here, $\sum_i f_i f_i^*$ can be interpreted as converging in the ultraweak topology\footnote{For positive elements~$(x_i)_{i \in I}$ of a W$^*$-algebra, convergence and value of the sum~$\sum_{i \in I} x_i$ coincides in the ultraweak and ultrastrong topology and, in fact, equals the supremum of the partial sums.} in the case where $\cat{C}(B,B)$ is a W$^*$-algebra. If this is not the case, what we mean by the condition $\sum_i f_i f_i^* < \infty$ is that there is a fixed element which upper bounds all partial sums $\sum_{i\in S} f_i f_i^*$ for finite $S\subseteq I$, or equivalently that the sum converges ultraweakly in the double dual W$^*$-algebra $\cat{C}(B,B)^{**}$.

The set~\eqref{AiB} is a vector space under componentwise addition and scalar multiplication, where closure under addition follows from $f_i g^*_i + g_i f^*_i \leq f_i f^*_i + g_i g^*_i$, which holds because of $(f_i - g_i) (f^*_i - g^*_i) \geq 0$. We furthermore consider this vector space as a normed space under the norm
\[
	\| (f_i)_{i \in I} \| \ :=\  \Bigl\| \textstyle \sum_i f_i f_i^* \Bigr\|^{1/2},
\]
where the triangle inequality again follows from $f_i g^*_i + g_i f^*_i \leq f_i f^*_i + g_i g^*_i$. We interpret the sum as before either in terms of a supremum over the norm of finite partial sums, or as converging in the double dual $\cat{C}(B,B)^{**}$.

@@

\begin{proof}
Suppose that we are given a Cauchy sequence of families $(f^{(n)})_{n\in\N} = ( ( f_i^{(n)} )_{i\in I} )_{n\in\N}$. For given $j$, the projection map
\[
	\bigoplus_i \cat{C}(A_i,B) \longrightarrow \cat{C}(A_j,B)	
\]
is bounded, and therefore $( f_j^{(n)} )_{n\in\N}$ is Cauchy as well; let us denote the limit by $f_j : A_j \to B$. We will now show that $f \in \bigoplus_i \cat{C}(A_i,B)$ for $f := (f_j)_{j\in I}$, which amounts to verifying the relevant square summability. The Cauchy assumption implies that the sequence $(f^{(n)})_{n\in\N}$ is uniformly bounded by some constant $C$,
\[
    \| f^{(n)} \| \ = \ 
	\Bigl\| \textstyle \sum_i f_i^{(n)} (f_i^{(n)})^* \Bigr\|^{1/2} \ \leq\  C.
\]
Restricting to a finite partial sum over $S\subseteq I$ and taking $n\to\infty$ shows that
\[
	\Bigl\| \textstyle \sum_{i\in S} f_i f_i^* \Bigr\|^{1/2} \ \leq\  C
\]
as well. Since $S$ was arbitrary, we conclude that $\| f \| \leq C$.

The convergence $f_i \to f$ works similarly. By assumption, for every~$\eps > 0$ we have~$\left\| f^{(n)} - f^{(m)} \right\| \leq \eps$ for almost all $n,m\in\N$. In particular, we know that for any finite $S \subseteq I$,
\[
	\Bigl\| \sum_{i\in S} \bigl(f_i^{(n)} - f_i^{(m)}\bigr) \bigl(f_i^{(n)} - f_i^{(m)}\bigr)^* \Bigr\|^{1/2} \ \leq\  \eps
\]
for almost all $n,m\in\N$. Taking $m\to\infty$ gives
\[
	\Bigl\| \sum_{i\in S} \bigl(f_i^{(n)} - f_i\bigr) \bigl(f_i^{(n)} - f_i\bigr)^* \Bigr\|^{1/2} \ \leq \ \eps.
\]
Since $S$ was arbitrary, we conclude $\| f^{(n)} - f \|  \leq \eps$ for almost all $n$.
\end{proof}
